---
title: "homework 8"
author: "JF"
date: "2025-03-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(AER)
library(readxl)
library(dplyr)
library(fixest)
library(ggplot2)
library(tidyverse)
library(car)
##########
cars <- read_excel("/Users/jadenfix/Desktop/Graduate School Materials/Advanced Econometrics 2/cars.xlsx")
auction1 <- read_excel("/Users/jadenfix/Desktop/Graduate School Materials/Advanced Econometrics 2/auction1.xlsx")
auction2 <- read_excel("/Users/jadenfix/Desktop/Graduate School Materials/Advanced Econometrics 2/auction2.xlsx")
```

A.)

OLS W FE
```{r}
cars <- cars |> 
  group_by(year, country) |> 
  mutate(
    market_share = qu / marketsize,  # Normalize within country-year
    total_inside_share = sum(market_share, na.rm = TRUE),
    outside_option = 1 - total_inside_share  # Common outside share for all products in a market
  ) |> 
  ungroup()

# Compute logit-transformed dependent variable
cars <- cars |> 
  mutate(y = log(market_share) - log(outside_option))

# Convert fixed effect variables to factors
cars <- cars |> 
  mutate(
    brand = as.factor(brand),
    country = as.factor(country),
    year = as.factor(year),
    log_pop = log(pop),
    log_gdp = log(ngdp)
  )

# Run OLS with Fixed Effects
ols_model <- feols(
  y ~ price + horsepower + fuel + width + height + weight + domestic + log_pop + log_gdp | brand + country + year, 
  data = cars, 
  vcov = "hc1"
)

# Show Results
summary(ols_model)
```
B.)
```{r}
cars <- cars |> 
  group_by(year, country) |> 
  mutate(
    # Compute market mean characteristic
    mean_horsepower = mean(horsepower, na.rm = TRUE),
    mean_fuel = mean(fuel, na.rm = TRUE),
    mean_width = mean(width, na.rm = TRUE),
    mean_height = mean(height, na.rm = TRUE),
    mean_weight = mean(weight, na.rm = TRUE),
    mean_domestic = mean(domestic, na.rm = TRUE)
  ) |> 
  mutate(
    # IV1: Difference from market mean
    iv_horsepower_1 = horsepower - mean_horsepower,
    iv_fuel_1 = fuel - mean_fuel,
    iv_width_1 = width - mean_width,
    iv_height_1 = height - mean_height,
    iv_weight_1 = weight - mean_weight,
    iv_domestic_1 = domestic - mean_domestic,

    # IV2: Squared difference from market mean
    iv_horsepower_2 = (horsepower - mean_horsepower)^2,
    iv_fuel_2 = (fuel - mean_fuel)^2,
    iv_width_2 = (width - mean_width)^2,
    iv_height_2 = (height - mean_height)^2,
    iv_weight_2 = (weight - mean_weight)^2,
    iv_domestic_2 = (domestic - mean_domestic)^2
  ) |> 
  ungroup()

# Run IV regression using feols() from fixest
iv_model <- feols(
  y ~ horsepower + fuel + width + height + weight + domestic + log_pop + log_gdp | brand + country + year | 
    price ~ iv_horsepower_1 + iv_horsepower_2 + iv_fuel_1 + iv_fuel_2 +
             iv_width_1 + iv_width_2 + iv_height_1 + iv_height_2 +
             iv_weight_1 + iv_weight_2 + iv_domestic_1 + iv_domestic_2, 
  data = cars, 
  vcov = "hc1"
)

# Show IV regression results
summary(iv_model)
```

```{r}
beta_fuel <- coef(iv_model)["fuel"]  
beta_price <- coef(iv_model)["fit_price"]  
# Convert price coefficient to Euros (price was in thousands of euros)  
willingness_to_pay <- (beta_fuel / beta_price) * 1000  
cat("Average willingness to pay for 1 liter per km improvement:", willingness_to_pay, "Euros\n")  
## Average willingness to pay for 1 liter per km improvement: -1467.287 Euros  

# Since fuel efficiency is liters per km, a decrease (improvement) is valued positively  
# Report absolute value  
cat("C: Average willingness to pay:", abs(willingness_to_pay), "Euros\n")  

# Test equality of country-specific price coefficients  
# Re-specify model with country-specific price interactions  
iv_model_countries <- feols(  
  y ~ price:country + horsepower + fuel + width + height + weight + domestic + log_pop + log_gdp |  
  brand + country + year |  
  price ~ iv_horsepower_1 + iv_horsepower_2 + iv_fuel_1 + iv_fuel_2 +  
          iv_width_1 + iv_width_2 + iv_height_1 + iv_height_2 +  
          iv_weight_1 + iv_weight_2 + iv_domestic_1 + iv_domestic_2,  
  data = cars,  
  vcov = "hc1"  
)  

# Wald test for equality of all country price coefficients  
wald_test_all <- linearHypothesis(iv_model_countries,  
                                  matchCoefs(iv_model_countries, "price:country"),  
                                  white.adjust = "hc1")  
# Print header first
cat("D: Wald Test all:\n")

# Print the test results
print(wald_test_all)  

# Wald test excluding UK (country5)  
countries_to_test <- levels(cars$country)[1:4] # Assuming country5 is UK  
wald_test_no_uk <- linearHypothesis(iv_model_countries,  
                                   paste0("price:country", countries_to_test, " = 0"),  
                                   white.adjust = "hc1")  
wald_test_no_uk
cat("D: Wald Test no UK:\n")
print(wald_test_no_uk)

# Correct elasticity calculation: alpha * price * (1 - market_share)  
cars <- cars |>  
  mutate(price_elasticity = coef(iv_model)["fit_price"] * price * (1 - market_share))  

# Summary statistics  
cat("E: price elasticity of demand for each product, market, and year:\n")
print(summary(cars$price_elasticity))  

```

2.)
In a first-price auction with three bidders and symmetric private values drawn from \[ F(x)=x^2,\quad x\in[0,1], \] the symmetric equilibrium bidding strategy, (b(v)), is given by \[ b(v) = v - \frac{\displaystyle \int_{0}^{v} \left(F(t)\right)^{n-1}\, dt}{\left(F(v)\right)^{n-1}}, \] where (n=3). Since \[ F(v)=v^2,\quad \text{so} \quad \left(F(v)\right)^{2} = \left(v^2\right)^2 = v^4, \] and \[ \int_0^v \left(F(t)\right)^{2}\, dt = \int_0^v (t^2)^2\, dt = \int_0^v t^4\, dt = \frac{v^5}{5}, \] substituting these expressions into the bidding strategy yields \[ b(v)= v - \frac{\frac{v^5}{5}}{v^4} = v - \frac{v}{5} = \frac{4}{5}v. \]

Thus, the optimal bidding strategy is \[ b(v)=\frac{4}{5}v. \]

3.)
a.)
```{r}
f_sealed_2bid <- function(bids, eps=0.5){
  values <- rep(NA, length(bids))
  for (i in 1:length(bids)){
    H_hat <- mean(bids < bids[i])
    h_hat <- (mean(bids < bids[i] + eps) - mean(bids < bids[i] - eps))/(2*eps)
    values[i] <- bids[i] + H_hat/h_hat
  }
  return(values)
}
bids <- auction1[2:3]
bids <- as.vector(as.matrix(bids))
values <- f_sealed_2bid(bids)

dv<- density(values)
db<- density(bids)
plot(db,col="blue", xlim=c(-5, 15), lwd = 5)
lines(dv,col="green", lwd= 3)
abline(h=3)
```
B.)
```{r}
bids <- auction1[, 2:3]
bids <- as.vector(as.matrix(bids))   
values <- f_sealed_2bid(bids)        
hist(bids, 
     col = rgb(0, 0, 1, 0.5), 
     breaks = 20, 
     xlim = c(0,50), 
     ylim = c(0, max(hist(bids, plot = FALSE)$counts, 
                     hist(values, plot = FALSE)$counts)), 
     xlab = "Bid/Value", 
     main = "Histogram of Observed Bids and Estimated Valuations")

hist(values, 
     col = rgb(0, 1, 0, 0.5),  
     breaks = 200, 
     add = TRUE)               
legend("topright", 
       legend = c("Bids", "Estimated Valuations"), 
       fill = c(rgb(0, 0, 1, 0.5), rgb(0, 1, 0, 0.5)), 
       border = "black")
```

```{r}
mean_bids <- mean(bids)
mean_vals <- mean(values)
med_bids <- median(bids)
med_vals <- median(values)
```
C.)
```{r}
neg_log_likelihood <- function(params, data) {
  mu <- params[1]  
  s <- params[2]  
  
  n <- length(data)
  log_likelihood <- -n/2 * log(2 * pi) - n * log(s) - sum((data - mu)^2) / (2 * s^2)
  
  return(-log_likelihood)  
}

init_params <- c(mean(values), sd(values))

mle_results <- optim(init_params, neg_log_likelihood, data = values, method = "L-BFGS-B",
                     lower = c(-Inf, 0.0001))
mu_hat <- mle_results$par[1]
s_hat <- mle_results$par[2]
cat("Estimated µ (mean):", mu_hat, "\n")
```

```{r}
cat("Estimated s (std dev):", s_hat, "\n")
```
4.)
```{r}
winning_bid <- auction2$Price

# Estimate the empirical CDF of the winning bid
G_emp <- ecdf(winning_bid)

# Create a grid of bid values spanning the range of winning bids
bid_grid <- seq(min(winning_bid), max(winning_bid), length.out = 200)

# Invert the relation: F(v) = 1 - sqrt(1 - G(v))
F_est <- 1 - sqrt(1 - G_emp(bid_grid))

# Assuming bid_grid and F_est are already defined

# Plot the estimated CDF
plot(bid_grid, F_est, type = "l", lwd = 4, col = "green",
     main = "Estimated Distribution of Valuations",
     xlab = "Valuation",
     ylab = "Estimated CDF")

# Overlay the normal CDF with mean and standard deviation estimated from data
mu_hat <- mean(bid_grid)  # Estimated mean
sigma_hat <- sd(bid_grid) # Estimated standard deviation

curve(pnorm(x, mean = mu_hat, sd = sigma_hat), 
      from = min(bid_grid), to = max(bid_grid), 
      col = "blue", lwd = 4, add = TRUE)

# Add legend
legend("bottomright", legend = c("Estimated CDF", "Normal CDF"),
       col = c("green", "blue"), lwd = 2)
```
